---
title: "36-726 Final Models"
author: "Melody Owen, Maitreyi Gunye, Emily Wang"
date: "April 8, 2018"
output: 
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---

```{r, setup, include=FALSE}
require(mosaic)   # Load additional packages here 
require(tidyr)
require(ggplot2)
require(car)
require(leaps)
require(MASS)
require(dplyr)
require(randomForest)

# Some customization.  You can alter or delete as desired (if you know what you are doing).
trellis.par.set(theme=theme.mosaic()) # change default color scheme for lattice
knitr::opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code

# Loading in Ethiopia datasets
ethiopia <- read.csv("../../Data/FinalData/ethiopiaFinal.csv",
                     stringsAsFactors = FALSE)
ethiopiaHousehold <- read.csv("../../Data/FinalData/ethiopiaHousehold.csv",
                     stringsAsFactors = FALSE)
ethiopiaIndividual <- read.csv("../../Data/FinalData/ethiopiaIndividual.csv",
                     stringsAsFactors = FALSE)

# Loading in Uganda datasets
uganda <- read.csv("../../Data/FinalData/ugandaFinal.csv",
                     stringsAsFactors = FALSE)
ugandaHousehold <- read.csv("../../Data/FinalData/ugandaHousehold.csv",
                     stringsAsFactors = FALSE)
ugandaIndividual <- read.csv("../../Data/FinalData/ugandaIndividual.csv",
                     stringsAsFactors = FALSE)
```

```{r}
# establishing cutoffs for the binary outcomes in Ethiopia
ethiopiaHouseholdm <- ethiopiaHousehold %>%
  mutate(childNumDefiniteHazardCat = ifelse(childNumDefiniteHazard > 0, 1, 0),
         childNumDefiniteWorstCat = ifelse(childNumDefiniteWorst > 0, 1, 0))

ethiopiaIndividualm <- ethiopiaIndividual %>%
  mutate(anyDefiniteHazardIndCat = ifelse(anyDefiniteHazardInd > 0, 1, 0),
         anyDefiniteWorstIndCat = ifelse(anyDefiniteWorstInd > 0, 1, 0))

# establishing cutoffs for the binary outcomes in Uganda
ugandaHouseholdm <-ugandaHousehold %>%
  mutate(childNumDefiniteHazardCat = ifelse(childNumDefiniteHazard > 0, 1, 0),
         childNumDefiniteWorstCat = ifelse(childNumDefiniteWorst > 0, 1, 0))

ugandaIndividualm <- ugandaIndividual %>%
  mutate(anyDefiniteHazardIndCat = ifelse(anyDefiniteHazardInd > 0, 1, 0),
         anyDefiniteWorstIndCat = ifelse(anyDefiniteWorstInd > 0, 1, 0))
```


```{r}
#set up training and testing sets in train dataset
set.seed(5)

#we seen before in the graphs there only a few observations of Age group less than 18. We delete this out of our data set.
set <- nrow(ethiopiaHouseholdm)*3/4
numtrain <- sample(1:dim(ethiopiaHouseholdm)[1], round(set), replace = FALSE, prob = NULL)
opp <- -numtrain
trainset <- ethiopiaHouseholdm[sort(numtrain), ]
testset <- ethiopiaHouseholdm[opp, ]
```


# Modeling for Ethiopia

##  Ethiopia Household Level Data

There are 2231 observations in the Ethiopia household data, meanining that there are 2231 different households.

### Predicting `childNumDefiniteHazard` and `childNumDefiniteHazardCat` in Ethiopia

In these models, we predict the number of children between ages 5 and 17 in a household that had any definite indication of being in hazardous work. We also predict this outcome as a binary outcome, i.e. if a house had any children in hazardous work.

```{r}
View(ethiopiaHouseholdm)
# childNumDefiniteHazard
fullModel <- lm(childNumDefiniteHazard ~ avg517Age + avgAdultAge + headType + 
                  totalChild517 + residenceType + homeRooms + totalIncome +
                  homeOwner + ageMax + ageMin,
                data = ethiopiaHouseholdm)
summary(fullModel)

reduce1 <- lm(childNumDefiniteHazard ~ avg517Age + avgAdultAge + headType + 
                  totalChild517 + residenceType + totalIncome +
                  homeOwner + ageMax + ageMin,
                data = ethiopiaHouseholdm)
summary(reduce1)
###
trainset <- ethiopiaHouseholdm[sort(numtrain), ]
testset 

###
model <- glm(Q8_Quality_of_Care ~.-PatientKey - ClientKey ,family=binomial(link='logit'),data=trainset)
print(summary(model), digits = 1)
AIC(model)
pred_glm <- predict(model, testset)
pred_glm_factor <- factor(as.numeric(pred_glm > 0), levels = 0:1 )
head(data.frame(pred_glm,pred_glm_factor), 30)
caret::confusionMatrix(pred_glm_factor, as.factor(testset$Q8_Quality_of_Care))

#coefficents for interpretation
data.frame(expcoef = round(exp(model$coefficients),2))

```


```{r}
# childNumDefiniteHazardCat
trainset <- ethiopiaHouseholdm[sort(numtrain), ]
levels(as.factor(testset$homeOwner))
levels(as.factor(trainset$homeOwner))
testset <- testset[!testset$homeOwner == "Subsidized by employer",]

fullModel <- glm(childNumDefiniteHazardCat ~ avg517Age + avgAdultAge + headType + 
                  totalChild517 + residenceType + homeRooms + totalIncome +
                  homeOwner + ageMax + ageMin,
                data = trainset, family = "binomial")
summary(fullModel)
print(summary(fullModel), digits = 1)
AIC(fullModel)
pred_glm <- predict(fullModel, testset[c(7,8,2,13,17,19,20,18,4,3)])
pred_glm_factor <- factor(as.numeric(pred_glm > 0), levels = 0:1 )
head(data.frame(pred_glm,pred_glm_factor), 30)
caret::confusionMatrix(pred_glm_factor, as.factor(testset$childNumDefiniteHazardCat))
data.frame(expcoef = round(exp(fullModel$coefficients),2))


fullModel1 <- glm(childNumDefiniteHazardCat ~ avg517Age +  headType + 
                  totalChild517 + residenceType +  totalIncome +
                   ageMin,
                data = trainset, family = "binomial")
summary(fullModel1)
print(summary(fullModel1), digits = 1)
AIC(fullModel1)
names(testset)
pred_glm1 <- predict(fullModel1, testset[,c(7,2,13,17,20,3)])
pred_glm_factor1 <- factor(as.numeric(pred_glm1 > 0), levels = 0:1 )
head(data.frame(pred_glm1,pred_glm_factor1), 30)
caret::confusionMatrix(pred_glm_factor, as.factor(testset$childNumDefiniteHazardCat))

data.frame(expcoef = round(exp(fullModel$coefficients),2))

anova(fullModel1,fullModel, test = "Chisq") 

Sensitivity <- data.frame(testset$childNumDefiniteHazardCat, pred_glm_factor1, testset$quesID)

#testset 0 then prediction be 1
Problems <- Sensitivity %>% filter(testset.childNumDefiniteHazardCat == 0, pred_glm_factor1 == 1) 
#find those problem rows in testset to look at
ProbX <- testset[testset$quesID %in% Problems$testset.quesID, ]
head(ProbX, 50)

prob <- ethiopiaHousehold[ethiopiaHousehold$quesID %in% Problems$testset.quesID, ]
View(prob)
sum(prob$hazardLaborHouseCat)/dim(prob)[1]

Problems <- Sensitivity %>% filter(testset.childNumDefiniteHazardCat == 0, pred_glm_factor1 == 0) 
#find those problem rows in testset to look at
ProbX <- testset[testset$quesID %in% Problems$testset.quesID, ]
head(ProbX, 50)

prob <- ethiopiaHousehold[ethiopiaHousehold$quesID %in% Problems$testset.quesID, ]
View(prob)
sum(prob$hazardLaborHouseCat)/dim(prob)[1]

#random Forest
#only does slightly better not worth it due to computational
trainset <- na.omit(trainset, cols = childNumDefiniteHazardCat) 
trainset=trainset %>% mutate_if(is.character, as.factor)
output.forest <- randomForest(childNumDefiniteHazardCat ~ avg517Age +  headType + 
                  totalChild517 + residenceType +  totalIncome +
                   ageMin, data = trainset, ntree=100, importance = TRUE)
importance <- importance(output.forest)
testset <- testset[c(7,2,13,17,20,18,3),]
names(testset)
pred_rf <- predict(output.forest, testset)
caret::confusionMatrix(pred_rf, testset$Q8_Quality_of_Care)
varImpPlot(output.forest, sort = T, n.var = 10, main = 'Top 10 Feature Importance')


plot(output.forest)

#use this plot to help us determine the number of trees. As the number of trees increases, 
#the OOB error rate decreases, and then becomes almost constant. We are not able to decrease
#the OOB error rate after about around 20 trees.
#want to tune random forest

responseCol <- match("Q8_Quality_of_Care", names(trainset))
extra <- match("PatientKey", names(trainset))
extra1 <- match("ClientKey", names(trainset))
t <- tuneRF(trainset[, -c(responseCol,extra)], trainset[, responseCol], stepFactor = 0.5, plot = TRUE, ntreeTry = 15, trace = TRUE, improve = 0.05)

#create 2nd random forest with tuned parameters
rfModel_new <- randomForest(Q8_Quality_of_Care ~ .- PatientKey -ClientKey, trainset, ntree = 15, mtry = 3, importance = TRUE, proximity = TRUE)
print(rfModel_new)
pred_rf2 <- predict(rfModel_new, testset)
caret::confusionMatrix(pred_rf2, testset$Q8_Quality_of_Care)
varImpPlot(rfModel_new, sort=T, n.var = 10, main = 'Top 10 Feature Importance')




```

### Predicting `childNumDefiniteWorst` and `childNumDefiniteWorstCat` in Ethiopia

In these models, we predict the number of children between ages 5 and 17 in a household that had any definite indication of being in the worst form of child labor. We also predict this outcome as a binary outcome.

```{r}
# childNumDefiniteWorst
fullModel <- lm(childNumDefiniteWorst ~ avg517Age + avgAdultAge + headType + 
                  totalChild517 + residenceType + homeRooms + totalIncome +
                  homeOwner + ageMax + ageMin,
                data = ethiopiaHouseholdm)
summary(fullModel)
```

```{r}
# childNumDefiniteWorstCat
fullModel <- glm(childNumDefiniteWorstCat ~ avg517Age + avgAdultAge + headType + 
                  totalChild517 + residenceType + homeRooms + totalIncome +
                  homeOwner + ageMax + ageMin,
                data = ethiopiaHouseholdm, family = "binomial")
summary(fullModel)
```

## Ethiopia Individual Level Data

There are 4251 observations in the Ethiopia individual level dataset, meaning that there are 4251 children between the ages of 5 and 17 in the data.

### Predicting `anyDefiniteHazardInd` and `anyDefiniteHazardIndCat` in Ethiopia

In these models, we predict the number of instances that a child between 5 and 17 answered "yes" for an indicator of hazardous work. We also predict this as a binary outcome, i.e. whether or not a child said yes to *any* indicator that would mean they are in hazarodous work. There are 31 total categories that contribute to this parameter.

```{r}
# anyDefiniteHazardInd
fullModel <- lm(anyDefiniteHazardInd ~ age + sex + readWrite +
                  headType + totalChild517 + residenceType + homeRooms + 
                  avgChildAge + avgAdultAge,
                data = ethiopiaIndividualm)
summary(fullModel)
```

```{r}
# anyDefiniteHazardIndCat
fullModel <- glm(anyDefiniteHazardIndCat ~ age + sex + readWrite +
                  headType + totalChild517 + residenceType + homeRooms + 
                  avgChildAge + avgAdultAge, 
                 data = ethiopiaIndividualm, family = "binomial")
summary(fullModel)
```

### Predicting `anyDefiniteWorstInd` and `anyDefiniteWorstIndCat` in Ethiopia

In these models, we predict the number of instances that a child between 5 and 17 answered "yes" for an indicator of worst child labor. We also predict this as a binary outcome, i.e. whether or not a child said yes to *any* indicator that would mean they are in the worst form of child labor. There are 37 total categories that contribute to this parameter.

```{r}
# anyDefiniteWorstInd
fullModel <- lm(anyDefiniteWorstInd ~ age + sex + readWrite +
                  headType + totalChild517 + residenceType + homeRooms + 
                  avgChildAge + avgAdultAge,
                data = ethiopiaIndividualm)
summary(fullModel)
```

```{r}
# anyDefiniteWorstIndCat
fullModel <- lm(anyDefiniteWorstIndCat ~ age + sex + readWrite +
                  headType + totalChild517 + residenceType + homeRooms + 
                  avgChildAge + avgAdultAge,
                data = ethiopiaIndividualm)
summary(fullModel)
```

# Modeling for Uganda

## Uganda Household Level Data

In the Uganda household level data, there are 617 observations, meaning 617 different households.

### Predicting `childNumDefiniteHazard` and `childNumDefiniteHazardCat` in Uganda

In these models, we predict the number of children between ages 5 and 17 in a household that had any definite indication of being in hazardous work. We also predict this outcome as a binary outcome, i.e. if a house had any children in hazardous work.

```{r}
# childNumDefiniteHazard
fullModel <- lm(childNumDefiniteHazard ~ ,
                data = ugandaHouseholdm)
summary(fullModel)
```

```{r}
# childNumDefiniteHazardCat
fullModel <- glm(childNumDefiniteHazardCat ~ ,
                data = ugandaHouseholdm, family = "binomial")
summary(fullModel)

```

### Predicting `childNumDefiniteWorst` and `childNumDefiniteWorstCat` in Uganda

In these models, we predict the number of children between ages 5 and 17 in a household that had any definite indication of being in the worst form of child labor. We also predict this outcome as a binary outcome. 

```{r}
# childNumDefiniteWorst
fullModel <- lm(childNumDefiniteWorst ~ ,
                data = ugandaHouseholdm)
summary(fullModel)
```

```{r}
# childNumDefiniteWorstCat
fullModel <- glm(childNumDefiniteWorstCat ~ ,
                data = ugandaHouseholdm, family = "binomial")
summary(fullModel)
```

## Uganda Individual Level Data

In the Uganda individual level data, there are 2264 observations, meaning there are 2264 children between the ages of 5 to 17.

### Predicting `anyDefiniteHazardInd` and `anyDefiniteHazardIndCat` in Uganda

In these models, we predict the number of instances that a child between 5 and 17 answered "yes" for an indicator of hazardous work. We also predict this as a binary outcome, i.e. whether or not a child said yes to *any* indicator that would mean they are in hazarodous work. There are 14 total categories that contribute to this parameter.

```{r}
# anyDefiniteHazardInd
fullModel <- lm(anyDefiniteHazardInd ~ ,
                data = ugandaIndividualm)
summary(fullModel)
```

```{r}
# anyDefiniteHazardIndCat
fullModel <- glm(anyDefiniteHazardIndCat ~ ,
                data = ugandaIndividualm, family = "binomial")
summary(fullModel)
```

### Predicting `anyDefiniteWorstInd` and `anyDefiniteWorstIndCat` in Uganda

In these models, we predict the number of instances that a child between 5 and 17 answered "yes" for an indicator of worst child labor. We also predict this as a binary outcome, i.e. whether or not a child said yes to *any* indicator that would mean they are in the worst form of child labor. There are 22 categories that contribute to this parameter.

```{r}
# anyDefiniteWorstInd
fullModel <- lm(anyDefiniteWorstInd ~ ,
                data = ugandaIndividualm)
summary(fullModel)
```

```{r}
# anyDefiniteWorstIndCat
fullModel <- glm(anyDefiniteWorstIndCat ~ ,
                data = ugandaIndividualm, family = "binomial")
summary(fullModel)
```


